---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

``` {r}
library(lattice)
options(echo = TRUE, scipen = 1, digits = 2)
```

## Loading and preprocessing the data

``` {r}
activityData <- read.csv('activity.csv', colClasses=c("numeric", "character", "numeric"))
activityData$date <- as.POSIXct(paste(activityData$date, " ", activityData$interval %/% 100, ":", activityData$interval %% 100, sep = ""))
```


## What is mean total number of steps taken per day?

``` {r}
totalStepsByDay <- aggregate(activityData$steps, by=list(as.Date(activityData$date)), FUN=sum, drop = TRUE)

hist(totalStepsByDay$x)
```

Mean steps taken by day is `r mean(totalStepsByDay$x, na.rm = TRUE)`, median is `r median(totalStepsByDay$x, na.rm = TRUE)`.

## What is the average daily activity pattern?

``` {r}
dailyActivityPattern <- aggregate(x = activityData$steps, by=list(activityData$interval), mean, na.rm=TRUE)
rownames(dailyActivityPattern) <- dailyActivityPattern$Group.1
maxTimeInterval <- dailyActivityPattern[order(-dailyActivityPattern$x)[1], ]
plot(dailyActivityPattern, type = "l")
```

Time interval with the most steps is `r maxTimeInterval$Group.1` with `r maxTimeInterval$x` steps.

## Imputing missing values

``` {r}
missingValues <- activityData[is.na(activityData$steps), ]
```

Total number of missing values in dataset is `r nrow(missingValues)` (`r (nrow(missingValues) / nrow(activityData)) * 100` %).

``` {r}
imputedDataSet <- activityData
imputedDataSet[is.na(activityData$steps), ]$steps <- unlist(lapply(missingValues$interval, function(x) { dailyActivityPattern[as.character(x), ]$x }))

totalStepsByDayImputed <- aggregate(imputedDataSet$steps, by=list(as.Date(activityData$date)), FUN=sum)

hist(totalStepsByDayImputed$x)
```

Mean steps taken by day in the imputed dataset is `r mean(totalStepsByDayImputed$x, na.rm = TRUE)`, median is `r median(totalStepsByDayImputed$x, na.rm = TRUE)`.

## Are there differences in activity patterns between weekdays and weekends?

``` {r}
imputedDataSet$weekday <- as.factor(unlist(lapply(weekdays(activityData$date), function(d) { if (d == "Saturday" | d == "Sunday") { "weekend" } else { "weekday" }})))
dailyActivityPatternPerWeekday <- aggregate(x = activityData$steps, by=list(imputedDataSet$interval, imputedDataSet$weekday), mean, na.rm=TRUE)

xyplot(x ~ Group.1 | Group.2, data=dailyActivityPatternPerWeekday, type="l", layout=c(1,2))
```